---
title: "Comparison of Recycling Systems in England"
subtitle: "Does the collection system have a meaningful impact on recycling rate?"
author: "Clare Gibson"
date: today
format:
  html:
    theme: [cosmo, custom.scss]
    toc: true
    toc-location: left
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    embed-resources: true
    fig-align: center
    include-in-header:
      text: |
        <link rel="preconnect" href="https://fonts.gstatic.com" crossorigin>
        <link href="https://fonts.googleapis.com/css2?family=Work+Sans:ital,wght@0,300;0,600;1,300&family=Roboto+Mono:wght@600&display=swap" rel="stylesheet">
execute:
  warning: false
  message: false
---

![](img/dt-logo.svg){.dt-logo fig-alt="Data Translator logo"}

# Introduction

This paper sets out to answer the question: **Does the collection system have a meaningful impact on recycling rate?**

Local authorities in England can elect to use one of three collection systems for household recycling:

- **Co-Mingled:** all non-organic recyclable material is put into a single bin by the householder
- **Two Stream:** non-organic recyclable material must be sorted into two bins by the householder (e.g. one bin for glass and metal, and a second bin for paper, card and plastic)
- **Multi-Stream:** householders sort their non-organic recyclable material into separate bins for each type of material

Multi-Stream systems place a higher burden on the householder, so this paper asks whether that higher burden is associated with higher recycling rates, and whether any difference is large enough to matter in practice.

The word "meaningful" in the research question is important. A difference in recycling rates can be statistically detectable without being practically meaningful. This analysis therefore reports not only whether the systems differ, but by how much, and with how much confidence.

A note on causation: local authorities choose their own collection systems. Therefore, this is observational data and not a randomised experiment. Throughout this paper the language is deliberately one of association rather than cause.

In this analysis, we set the threshold for statistical significance at *p* < 0.05.

# Setup
## Load required packages
```{r load-packages}
library(here)
library(tidyverse)
library(knitr)
library(moments)
library(rstatix)
library(broom)
library(showtext)
```

``` {r source-code}
source(here("R/theme_datatranslator.R"))
```

``` {r set-typography}
font_add_google("Work Sans", "Work Sans")
font_add_google("Roboto Mono", "Roboto Mono")
showtext_auto()
showtext_opts(dpi = 96)
```

## Import the recycling data

```{r import-data}
recycling <- 
  read_csv(here("data/cln/data_recycling_system.csv")) |> 
  # reorder factor levels to assist with plotting
  mutate(
    collection_type = fct_reorder(collection_type, household_recycle_rate)
  )

head(recycling) |> kable()
```

The dataset includes information about household recycling for `r nrow(recycling)` local authorities in England in 2021. It contains the unique code and name of each authority, the population of the authority in 2021, the type of collection system in place in 2021, the amount of household waste collected and recycled in 2021 (in tonnes) and the household recycling rate.

## Helper code

The chunk below contains code to support consistent plotting.

```{r plotting-summarising-code}

# We'll be creating a number of similar plots, so we can wrap the code into a function
plot_distribution <- function(data, yvar, title, scale = c("none", "percent", "thousands")) {
  scale <- match.arg(scale)

  p <- ggplot(data, aes(x = collection_type, y = {{ yvar }})) +
    geom_violin(aes(fill = collection_type, colour = collection_type),
                linewidth = 1, alpha = 0.5) +
    geom_boxplot(outlier.alpha = 0, coef = 0,
                 colour = dt_navy, width = 0.2) +
    stat_summary(fun = mean, geom = "point", shape = 18, size = 3,
                 colour = dt_navy) +
    coord_flip() +
    scale_colour_dt(guide = "none") +
    scale_fill_dt(guide = "none") +
    labs(x = NULL, y = NULL, title = title) +
    theme_dt() +
    theme(
      panel.grid.major.y = element_blank(),
      panel.grid.minor   = element_blank(),
      plot.background = element_rect(fill = "#ECF0F4")
    )

  if (scale == "percent") {
    p <- p + scale_y_continuous(labels = scales::percent)
  }
  if (scale == "thousands") {
    p <- p + scale_y_continuous(labels = scales::label_number(scale = 1e-3))
  }

  p
}

# Helper to set format and structure of stats summaries
summarise_distribution <- function(data, yvar) {
  data |>
    group_by(collection_type) |>
    summarise(
      n      = n(),
      mean   = mean({{ yvar }}, na.rm = TRUE),
      median = median({{ yvar }}, na.rm = TRUE),
      sd     = sd({{ yvar }}, na.rm = TRUE),
      skew   = skewness({{ yvar }})
    )
}
```

The chunk below contains code to support reporting of results.

```{r reporting-helpers}
# Convert a proportion to percentage points
pp <- function(x, digits = 1) round(x * 100, digits)

# Count and share of LAs using a given collection system
n_sys   <- function(x) sum(recycling$collection_type == x)
pct_sys <- function(x) round(mean(recycling$collection_type == x) * 100)

# Pull a single value out of a broom tidy tibble by term name
get_term <- function(tbl, term_name, col = "estimate") {
  tbl[[col]][tbl$term == term_name]
}
```

# Visualise the distributions
Before we run a significance test, let's view both the number of local authorities employing each type of collection system (see @fig-collection-systems), and the distributions of recycling rates by collection type (see @fig-recycle-rate).

## Breakdown of collection systems
```{r}
#| label: fig-collection-systems
#| fig-cap: "Count of local authorities by collection system. Group sizes are markedly uneven, which informs the choice of statistical test."

ggplot(recycling, aes(collection_type, fill = collection_type)) +
geom_bar() +
geom_text(stat = "count", aes(label = after_stat(count)), 
          vjust = -0.5, family = "Work Sans", colour = dt_navy) +
scale_fill_dt(guide = "none") +
scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
labs(
  x = NULL,
  y = NULL,
  title = "Count of local authorities by recycling collection system"
) +
theme_dt() +
theme(
  axis.text.y  = element_blank(),
  axis.ticks.y = element_blank(),
  plot.background = element_rect(fill = "#ECF0F4"),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank()
)
```

The bar chart in @fig-collection-systems shows that the group sizes are uneven. `r n_sys("Co-Mingled")` (`r pct_sys("Co-Mingled")`%) authorities use a Co-Mingled system, `r n_sys("Two Stream")` (`r pct_sys("Two Stream")`%) use a Two Stream system and `r n_sys("Multi-Stream")` (`r pct_sys("Multi-Stream")`%) use a Multi-Stream system.

## Distribution of recycling rates by system
```{r}
#| label: fig-recycle-rate
#| fig-cap: "Distribution of household recycling rates by collection system. The box shows the median and interquartile range; the diamond shows the mean."

plot_distribution(
  recycling,
  household_recycle_rate,
  title = "Distribution of recycling rates by collection system",
  scale = "percent"
)
```

```{r}
#| label: tbl-recycle-rate
#| tbl-cap: "Summary statistics for household recycling rate by collection system"

summarise_distribution(recycling, household_recycle_rate) |> kable(digits = 3)
```

As shown in @tbl-recycle-rate, we can see that Multi-Stream collection systems achieved the highest mean and median recycling rate in 2021, and Co-Mingled systems achieved the lowest rates. The skew number tells us that the distributions are not normal. Co-Mingled systems show a positive skew, while Two Stream and Multi-Stream both show a negative skew.

## Why we analyse rate rather than volume

The raw volumes of household waste recycled by local authority are heavily right-skewed, as shown in the plots below. They are dominated by local authority size, since a large council collects and recycles more of everything. This is why the analysis works on the rate of recycling, which normalises for size. It is also why we will use `log(population)` when we come to control for size in the regression.

```{r}
#| label: fig-household-collected
#| fig-cap: "Distribution of household waste collected by collection system. Values are in thousands of tonnes. The box shows the median and interquartile range; the diamond shows the mean."

plot_distribution(
  recycling,
  household_collected,
  title = "Distribution of household waste collected by collection system",
  scale = "thousands"
)
```

```{r}
#| label: tbl-household-collected
#| tbl-cap: "Summary statistics for amount of household waste collected (tonnes) by collection system"

summarise_distribution(recycling, household_collected) |> kable(digits = 3)
```

```{r}
#| label: fig-household-recycled
#| fig-cap: "Distribution of household waste recycled by collection system. Values are in thousands of tonnes. The box shows the median and interquartile range; the diamond shows the mean."

plot_distribution(
  recycling,
  household_recycled,
  title = "Distribution of household waste recycled by collection system",
  scale = "thousands"
)
```

```{r}
#| label: tbl-household-recycled
#| tbl-cap: "Summary statistics for amount of waste recycled (in tonnes) by collection system"

summarise_distribution(recycling, household_recycled) |> kable(digits = 3)
```

# Test for significance

## Kruskal-Wallis test

The default statistical test would be a one-way ANOVA, which compares group means. But ANOVA assumes each group is roughly normally distributed with similar variance, and our groups are uneven (`r n_sys("Co-Mingled")` / `r n_sys("Two Stream")` / `r n_sys("Multi-Stream")`) and somewhat skewed, as shown above. Instead, we will use the Kruskal-Wallis test: the non-parametric equivalent. It works on the ranks of the values instead of the raw values, so it makes no assumption about the shape of the distribution, and it handles unequal group sizes comfortably. It tests whether at least one group tends to sit higher than the others.

```{r kruskal-wallis-test}
# Run the KW test for recycling rate as explained by collection type
kw <- kruskal.test(household_recycle_rate ~ collection_type, data = recycling)
kw
```

The test returns H = `r round(kw$statistic, 2)` on `r kw$parameter` degrees of freedom, *p* = `r round(kw$p.value, 3)`. At least one collection system therefore differs from the others to a statistically significant degree.

This tells us that a difference exists somewhere, but not where, and not how large. Both of those questions matter more than the first.

## Effect size

However, a *p*-value only tells us whether there is an effect, not how big it is. With `r nrow(recycling)` authorities, even a trivial difference can come out significant. Our question asks about a meaningful impact, so we need an effect size, which measures how much of the variation in recycling rates is explained by collection system.

```{r}
#| label: tbl-kw-effect
#| tbl-cap: "Kruskal-Wallis effect size for recycling rates as explained by collection system"


eff <- recycling |> kruskal_effsize(household_recycle_rate ~ collection_type)
eff |> kable(digits = 3)
```

Eta-squared is `r round(eff$effsize, 3)`, meaning collection system accounts for `r round(eff$effsize * 100, 1)`% of the variation in recycling rates between local authorities. By the conventional thresholds (0.01 small, 0.06 medium, 0.14 large) this is a small effect. This means that the collection system is not the biggest contributor to the recycling rate.

## Dunn's test

Finally, we can run Dunn's test to test the collection systems in a pairwise manner, which allows us to see whether any one of them show a significant difference in recycling rate compared to the others.

```{r}
#| label: tbl-dunns-test
#| tbl-cap: "Summary of Dunn's test results"


medians <- recycling |> 
  group_by(collection_type) |> 
  summarise(median_rate = median(household_recycle_rate)) |> 
  mutate(collection_type = as.character(collection_type))

dunn <- recycling |> 
  dunn_test(household_recycle_rate ~ collection_type,
            p.adjust.method = "holm") |> 
  left_join(medians, by = join_by(group1 == collection_type)) |> 
  rename(median1 = median_rate) |> 
  left_join(medians, by = join_by(group2 == collection_type)) |> 
  rename(median2 = median_rate) |> 
  mutate(median_gap_pp = pp(median2 - median1)) |> 
  select(-.y., -median1, -median2)

# Function to pull out data from Dunn test
dunn_gap  <- function(g1, g2) dunn$median_gap_pp[dunn$group1 == g1 & dunn$group2 == g2]
dunn_padj <- function(g1, g2) dunn$p.adj[dunn$group1 == g1 & dunn$group2 == g2]
dunn_p <- function(g1, g2) dunn$p[dunn$group1 == g1 & dunn$group2 == g2]

dunn |> kable(digits = 3)
```

The Kruskal-Wallis test is an omnibus test: it reports that some difference exists without identifying which systems differ. Dunn's test makes the three pairwise comparisons, using the same pooled ranks. Because running three tests inflates the chance of a false positive, the *p*-values are adjusted using the Holm correction, and it is the adjusted values (`p.adj` in the table above) that should be interpreted.

Only one pair differs significantly after correction. Co-Mingled authorities are associated with lower recycling rates than Multi-Stream authorities (`p.adj` = `r round(dunn_padj("Co-Mingled", "Multi-Stream"), 3)`), a median gap of `r round(dunn_gap("Co-Mingled", "Multi-Stream"),1)`pp.

The other two comparisons are more nuanced. The Co-Mingled versus Two Stream comparison narrowly misses our stated
threshold for significance (`p.adj` = `r round(dunn_padj("Co-Mingled","Two Stream"), 3)`), and yet the median gap is `r round(dunn_gap("Co-Mingled","Two Stream"),1)`pp, which is comparable in size to the significant result above. The evidence here is suggestive but not conclusive.

The Two Stream versus Multi-Stream comparison is different. It is not significant (`p.adj` = `r round(dunn_padj("Two Stream","Multi-Stream"), 3)`) and the median gap is a negligible `r round(dunn_gap("Two Stream","Multi-Stream"),1)`pp. Here both the test and the estimate point the same way, towards there being little or no real difference between the two.

Taken together, the pattern is consistent with the burden of separation mattering mainly at the first step, going from mixing everything into one bin to separation into two bins, with little further gain from separating into more streams. The data support this interpretation but do not establish it. With only `r n_sys("Multi-Stream")` Multi-Stream authorities, this analysis has limited power to detect modest differences.

# Controlling for authority size

Significance testing has now been taken about as far as it can usefully go. Every result has been interpreted as either significant or not significant, with one landing almost exactly on the threshold. We now have two further questions to answer.

Firstly, how large is each difference, and how certain are we? We can answer this question with regression by reporting effect estimates with confidence intervals, which say more than a *p*-value alone.

Secondly, is the collection system really doing the work? If Multi-Stream authorities happen to be systematically smaller, then what looks like a system effect may be a size effect in disguise. This is an example of a confounding variable. Regression addresses this by comparing systems among authorities of similar size, holding size constant.

Population is used on a log scale. As the volume distributions in @fig-household-collected and @fig-household-recycled showed, authority size is heavily right-skewed, with a long tail of very large councils. Logging the variable means the model reads proportional differences in size, such as a doubling of population, rather than treating each additional resident as equivalent whether it is the two-thousandth or the millionth.

Co-Mingled is the reference level, so each coefficient represents the difference in recycling rate relative to a Co-Mingled authority. Because the recycling rate is a proportion, a coefficient of 0.04 corresponds to 4pp.

```{r regression-analysis}
mod_1 <- lm(household_recycle_rate ~ collection_type, data = recycling)
mod_2 <- lm(household_recycle_rate ~ collection_type + log(population), data = recycling)

tidy_1   <- tidy(mod_1, conf.int = TRUE)
tidy_2   <- tidy(mod_2, conf.int = TRUE)
glance_1 <- glance(mod_1)
glance_2 <- glance(mod_2)
```

```{r}
#| label: tbl-regression-tidy-2
#| tbl-cap: "Summary of regression for data controlled for population"


tidy_2 |> kable(digits = 4)
```

```{r}
#| label: tbl-regression-glance-2
#| tbl-cap: "Further summary of regression for data controlled for population"


glance_2 |> kable(digits = 4)
```

```{r}
#| label: tbl-regression-tidy-1
#| tbl-cap: "Summary of regression for data uncontrolled for population"

tidy_1 |> kable(digits = 4)
```

```{r}
#| label: tbl-regression-glance-1
#| tbl-cap: "Further summary of regression for data uncontrolled for population"

glance_1 |> kable(digits = 4)
```

## Interpreting the regression models

The regression analysis showed two results.

Firstly, the system effect was robust to authority size. The collection type coefficients barely changed when population was factored in: Two Stream moved from `r pp(get_term(tidy_1, "collection_typeTwo Stream"))`pp to `r pp(get_term(tidy_2, "collection_typeTwo Stream"))`pp, and Multi-Stream from `r pp(get_term(tidy_1, "collection_typeMulti-Stream"))`pp to `r pp(get_term(tidy_2, "collection_typeMulti-Stream"))`pp. If size had been confounding the comparison, these would have reduced materially. The association between collection system and recycling rate is not an artefact of larger or smaller councils favouring particular systems.

Secondly, larger authorities recycle less. The coefficient on `log(population)` is `r round(get_term(tidy_2, "log(population)"), 3)` (*p* = `r round(get_term(tidy_2, "log(population)", "p.value"), 3)`), which corresponds to a fall of roughly `r abs(pp(get_term(tidy_2, "log(population)") * log(2)))`pp in the recycling rate for each doubling of population. This is a modest but real association, and it is independent of the collection system in use.

The confidence intervals are more informative than the *p*-values. Relative to Co-Mingled, Two Stream is associated with an increase of `r pp(get_term(tidy_2, "collection_typeTwo Stream"))`pp (95% CI: `r pp(get_term(tidy_2, "collection_typeTwo Stream", "conf.low"))`pp to `r pp(get_term(tidy_2, "collection_typeTwo Stream", "conf.high"))`pp) and Multi-Stream with an increase of `r pp(get_term(tidy_2, "collection_typeMulti-Stream"))`pp (95% CI: `r pp(get_term(tidy_2, "collection_typeMulti-Stream", "conf.low"))`pp to `r pp(get_term(tidy_2, "collection_typeMulti-Stream", "conf.high"))`pp). Both intervals exclude zero, so both effects are likely real, but both are wide. The Multi-Stream interval spans everything from a barely noticeable `r round(pp(get_term(tidy_2, "collection_typeMulti-Stream", "conf.low")))`pp to a substantial `r round(pp(get_term(tidy_2, "collection_typeMulti-Stream", "conf.high")))`pp. A fair interpretation of this is that a real effect exists and its likely size is modest, but this dataset cannot pin it down precisely.

There is one apparent discrepancy that we should explain. In the regression, Two Stream differs significantly from Co-Mingled (*p* = `r round(get_term(tidy_2, "collection_typeTwo Stream", "p.value"),3)`), whereas Dunn's test found the same comparison borderline (`p.adj` = `r round(dunn_padj("Co-Mingled","Two Stream"),3)`). These are not contradictory. Dunn's raw *p*-value for that pair was `r round(dunn_p("Co-Mingled","Two Stream"),3)`, and the Holm correction for three comparisons roughly doubled it. The regression applies no such correction. The two results agree on the substance: there is a difference of around 3 to 4pp, and the evidence for it is real but not overwhelming.

Finally, the model explains very little. The R-squared is `r round(glance_2$r.squared, 3)`. This means that `r round((1 - glance_2$r.squared) * 100)`% of the variation in recycling rates between English local authorities lies outside collection system and population entirely.

# Conclusion

**Does the collection system have a meaningful impact on recycling rates?**

The answer has three parts, and they pull in different directions.

**Yes, there is a real difference, and it is not trivial in size.** Authorities that require householders to separate their recycling are associated with higher recycling rates than those that collect everything co-mingled. Adjusting for authority size, the gap is `r pp(get_term(tidy_2, "collection_typeTwo Stream"))`pp for Two Stream and `r pp(get_term(tidy_2, "collection_typeMulti-Stream"))`pp for Multi-Stream, relative to Co-Mingled. A 4pp shift in the national recycling rate would be a substantial policy outcome, so this is not a difference that can be dismissed as
academic.

**But the effect appears to be a step, not a ladder.** The evidence does not support the intuition that asking householders to sort into ever more streams yields ever higher rates. Two Stream and Multi-Stream authorities are statistically indistinguishable from one another, with a median gap of `r round(dunn_gap("Two Stream","Multi-Stream"),1)`pp. What appears to matter is whether households separate their recycling at all, not how finely they separate it. If this interpretation is correct, it has a clear practical implication: the additional burden that Multi-Stream places on householders, relative to Two Stream, gives no detectable improvement. This is the most interesting finding in the analysis, and it is also the one held most tentatively, resting as it does on only `r n_sys("Multi-Stream")` Multi-Stream authorities.

**And most of what drives recycling rates lies elsewhere.** Collection system alone explains `r round(glance_1$r.squared * 100, 1)`% of the variation between authorities. Adding population raises this only to `r round(glance_2$r.squared * 100, 1)`%. The other `r 100-round(glance_2$r.squared * 100, 1)`% belongs to factors this analysis cannot see. A local authority hoping to transform its recycling performance by changing its bin system alone would, on this evidence, be disappointed. The system is one lever among many, and not the largest.

There are also sound reasons why an authority might choose a system irrespective of its effect on rates. Without access to a Materials Recovery Facility, where co-mingled waste can be mechanically sorted, a council may have little practical alternative to kerbside sorting. The choice of system is constrained by infrastructure, cost and geography, not made freely on the basis of expected recycling rate.

## Data sources and limitations

1. **Observational, not experimental.** Local authorities chose their own collection systems. Any of the associations reported here could reflect the kind of authority that adopts a given system rather than the system itself. Nothing in this analysis establishes cause.

2. **Residual confounding is likely.** The only control available in this dataset is population. The factors that the recycling literature would flag most strongly are absent: deprivation, rural versus urban character, and housing stock. Housing type in particular is a known driver, since flats recycle markedly less well than houses whatever bins are provided, and it is plausibly correlated with the choice of system. The regression should therefore be read as adjusting for authority size only.

3. **A single year.** All figures relate to 2021. Rates fluctuate, and 2021 was likely affected by altered patterns of household waste generation during the pandemic. A multi-year panel would be more robust and would allow authorities that changed system to be observed before and after, which is a considerably stronger design.

4. **Unequal and small groups.** Only `r n_sys("Multi-Stream")` of the `r nrow(recycling)` authorities operate Multi-Stream systems. Comparisons involving this group have limited statistical power, and the finding that Two Stream and Multi-Stream cannot be distinguished may reflect that limitation rather than a true absence of difference.

5. **System classification.** Collection system data was obtained by Freedom of Information request and reflects the system in place at the time of response. Authorities operating hybrid or transitional arrangements are assigned to a single category, which introduces some measurement error.

## Further work

- Test Two Stream against Multi-Stream directly by releveling the factor, rather than inferring it from two comparisons against a shared reference.
- Join contextual variables (deprivation, rural or urban classification, housing type) and re-fit the model. If the system effect survives these controls, the finding is considerably stronger.
- Use those contextual variables to construct peer groups of comparable authorities through cluster analysis, deliberately excluding both the recycling rate and the collection system, and then compare systems within peer groups. This controls for confounding by stratification.
- Extend to multiple years to exploit authorities that changed system.