---
title: "A cluster analysis of recycling systems in England"
author: "Clare Gibson"
output:
  html_document:
    theme:
      bg: "#ECF0F4"
      fg: "#202C39"
      primary: "#FF4F5C"
      base_font:
        google: "Figtree"
      heading_font:
        google: "Oswald"
      code_font:
        google: "Roboto Mono"
    toc: true
    toc_float: true
    toc_depth: 2
---

# Setup
## Packages
Let's load the required packages for this analysis.
```{r load-packages, message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(skimr)
library(corrplot)
library(tidytable)
library(janitor)
library(knitr)
```

## Collect the data
Let's import the data and preview it.
```{r import-data, message=FALSE, warning=FALSE}
recycling <- 
  read_csv(here("data/cln/data.csv"))

glimpse(recycling)
```

The dataset contains a large number of character columns. Let's select only the columns we are interested in for this analysis.
```{r select-columns}
recycling <- 
  recycling |> 
  select(
    lad21cd,
    population,
    collection_type,
    household_collected,
    household_recycled,
    household_not_recycled,
    household_rejects
  )

head(recycling)
```

Now we can check the summary statistics for each column and see if we have any missing values.

```{r summary-stats}
skim(recycling)
```

We can see that there are no missing values in any of the columns. We notice that `household_rejects` has a some negative values, which doesn't make sense. I propose to ignore that column for my analysis.

# Exploratory Analysis
## Correlations
Let's run a correlation matrix to understand relationships between the variables.
```{r correlation-matrix}
recycling |> 
  select(where(is.numeric)) |> 
  cor(use = "pairwise.complete.obs") |> 
  round(2) |> 
  corrplot(method = "color", type = "upper")
```

There is a high correlation between `population` and `household_collected` meaning that as the population increases, so does the total amount of household waste that is collected. Same goes for `population` vs `household_not_recycled`. Both of these seem fairly obvious, so what if we look at the proportion of waste that is recycled instead of the amount? To do this we need to engineer a new feature into the dataset.

```{r add-household_recycled_pct}
recycling <- 
  recycling |> 
  mutate(
    household_recycled_pct = round(household_recycled / household_collected, 3)
  )

recycling |> 
  select(where(is.numeric)) |> 
  cor(use = "pairwise.complete.obs") |> 
  round(2) |> 
  corrplot(method = "color", type = "upper")
```

Now we see a negative correlation between `population` and `household_recycled_pct`, meaning that as population increases, the proportion of waste that is recycled decreases. Larger populations are associated with lower recycling rates.

Let's explore this relationship in a scatterplot.
```{r scatter-population-recycle-rate}
recycling |> 
  ggplot(
    aes(
      x = household_recycled_pct,
      y = population
    )
  ) +
  geom_point() +
  geom_smooth()
```

# Clustering
## Normalization
```{r normalize-data}
scaled_recycling <- 
  recycling |> 
  select(where(is.numeric)) |> 
  scale()
```

## Fit the model
```{r fit-the-km-model}
# Get the 2 columns of interest
scaled_recycling_2col <- 
  scaled_recycling[, c("population", "household_recycled_pct")]

set.seed(123)
km.out <- kmeans(scaled_recycling_2col, centers = 3, nstart = 20)
km.out
```

## Optimize the model
```{r optimize-model}
# Decide how many clusters to look at
n_clusters <- 10

# Initialize total within sum of squares error: wss
wss <- numeric(n_clusters)

set.seed(123)

# Look over 1 to n possible clusters
for (i in 1:n_clusters) {
  # Fit the model
  km.out <- kmeans(scaled_recycling_2col, centers = i, nstart = 20)
  # Save the within cluster sum of squares
  wss[i] <- km.out$tot.withinss
}

# Produce a scree plot
wss_df <- tibble(clusters = 1:n_clusters, wss = wss)

scree_plot <- ggplot(wss_df, aes(x = clusters, y = wss, group = 1)) +
  geom_point(size = 4) +
  geom_line() +
  scale_x_continuous(breaks = c(2, 4, 6, 8, 10)) +
  xlab("Number of clusters")

scree_plot
```

## Rebuild the model with optimized k
```{r optimized-k}
k <- 6

set.seed(123)

km.out <- kmeans(scaled_recycling_2col, centers = k, nstart = 20)
```

```{r scatterplot-with-clusters}
recycling$cluster_id <- factor(km.out$cluster)

ggplot(
  recycling, 
  aes(
    x = population,
    y = household_recycled_pct,
    colour = cluster_id
  )
) +
  geom_point() +
  xlab("Population") +
  ylab("Recycling Rate")
```

# Interpret the results
First, we dummy code the only categorical feature, `collection_type` to allow us to get descriptive statistics for the values of this feature.
```{r get-dummies}
# Get dummies
recycling <- 
  recycling |> 
  get_dummies(cols = collection_type) |> 
  select(!collection_type) |> 
  clean_names()

# Summarise descriptive stats of full data
recycling_stats <- 
  recycling |> 
  summarise(
    global_comingled = mean(collection_type_co_mingled),
    global_two_stream = mean(collection_type_two_stream),
    global_multi_stream = mean(collection_type_multi_stream),
    global_population = median(population),
    global_household_collected = median(household_collected),
    global_household_recycled = median(household_recycled),
    global_household_rate = median(household_recycled_pct)
  ) |> 
  round(2)

recycling_stats |> 
  kable()
```

Next get the same descriptive stats but grouped by cluster.
```{r stats-by-cluster}
recycling_stats_by_cluster <- 
  recycling |> 
  group_by(cluster_id) |> 
  summarise(
    cluster_comingled = round(mean(collection_type_co_mingled), 2),
    cluster_two_stream = round(mean(collection_type_two_stream), 2),
    cluster_multi_stream = round(mean(collection_type_multi_stream), 2),
    cluster_population = median(population),
    cluster_household_collected = median(household_collected),
    cluster_household_recycled = median(household_recycled),
    cluster_household_rate = median(household_recycled_pct)
  )

recycling_stats_by_cluster |> 
  kable()
```

Let's plot the differences from the population for each cluster.
```{r plot-differences}
recycling_cluster_profiles <- 
  recycling_stats_by_cluster |> 
  bind_cols(recycling_stats) |> 
  # Calculate the difference from global for each cluster
  mutate(
    type_comingled = (cluster_comingled - global_comingled) / global_comingled,
    type_two_stream = (cluster_two_stream - global_two_stream) / global_two_stream,
    type_multi_stream = (cluster_multi_stream - global_multi_stream) / global_multi_stream,
    population = (cluster_population - global_population) / global_population,
    household_collected = (cluster_household_collected - global_household_collected) / global_household_collected,
    household_recycled = (cluster_household_recycled - global_household_recycled) / global_household_recycled,
    household_rate = (cluster_household_rate - global_household_rate) / global_household_rate
  ) |> 
  select(cluster_id, type_comingled, type_two_stream, type_multi_stream, population,
         household_collected, household_recycled, household_rate) |> 
  pivot_longer(cols = !cluster_id, names_to = "feature", values_to = "percent_difference") |> 
  mutate(value_type = if_else(percent_difference >= 0, "positive", "negative"))

recycling_cluster_profiles |> 
  ggplot(aes(x = percent_difference, y = feature, fill = value_type)) +
  geom_col() +
  facet_wrap(~cluster_id, scales = "fixed")
```

